Meta-analytic approaches and effect sizes to account for 'nuisance heterogenity' in comparative physiology

### 2021-07-21 MS/ms.Rmd
Sammy's comments integrated

  pacman::p_load(flextable, tidyverse, equatags, png, magick)  
```

@@

Meta-analysis is always concerned with effect heterogeneity; in other words, the factors that drive differences in the direction and magnitude of effects within and across studies [@Nakagawa2017; @Gurevitch2018; @Cooperetal2019; @Borenstein2019]. The concept of heterogeneity is vitally important because it tells us how general findings are likely to be and how much of the variance we see is the result of real biology or methodological differences (not just sampling variance) [@Nakagawa2017; @Borenstein2019]. There are a plethora of phenomena driving effect heterogeneity, some of which are interesting and others that are less so.

@@

```{r, fig1, fig.cap="*Sources of effect size heterogeneity in meta-analyses*. Heterogeneity can come in various forms, from effect measures varying as a result of shared evolutionary history (i.e., phylogeny), different trait types measured, and sources of 'nuisance heterogeneity'. In this example, nuisance heterogeneity is generated by differences in the experimental temperatures at which the animals were tested. However, nuisance heterogeneity can come in various other forms such as differences in dosage, or biochemical variables between used in the experiments. We can control for these types of heterogeniety by reformalising effect sizes and/or using multilevel metaregression as we describe in our review"}
image <- image_read(here::here("figs", "Fig1.png"))
image
```

@@

We tend to agree with @Glass2015, and his sentiment is of particular relevance to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms, including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are different. By synthesizing diverse study outcomes, not only does it allow to report on overall effects (if the question is relevant to ask), but it allows us to also understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction) and approaches that can explain the diversity of effect outcomes observed. Doing so provides a richer set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included in a meta-analysis, and how this can affect interpretations made -- an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. If the estimate of an overall effect coincides with low heterogeneity estimates then the effects are reasonably consistent [@Borenstein2019]. If, however, an overall effect is accompanied by large heterogeneity estimates, then we would shift focus and explore drivers of that heterogeneity. For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant, and indeed may even be misleading. Therefore, understanding overall effects in each of these groups separately, and assess the extent to which they differ, may be more useful.

@@

Before embarking overviewing strategies to deal with 'nuisance heterogeneity', it is worth first understanding how comparative physiologists are currently conducting meta-analyses and other quantitative syntheses. As discussed above, different types of questions and experimental approaches will mix 'apples and oranges' to varying extents, but this heterogeneity (or basket of fruit) is often the main interest to comparative biologists. We conducted a literature survey to better understand the different types of effect sizes being used, their susceptibility to known sources of methodological / environmental variability across studies, and how these sources are controlled we conducted a literature survey. To accomplish this goal, we searched for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to continuous variation in how treatments were applied (e.g., for contrast-based effects) or the conditions under which traits were measured (i.e., for correlations) across studies (i.e., 'nuisance heterogeneity')? 3) If so, was the variation accounted for in the effect size or during analysis? and finally, 4) how many analyses used 'weighted' meta-analytic models?

@@

We included comparative / meta-analyses that were focused on comparative physiological questions. Narrative reviews and comments were not considered because they did not collate quantitative data from the literature. Studies needed to have collected physiological trait data from the literature such as hormones, thermal physiology (CTmax, CTmin, thermal preferences, thermal breadth, LT50, lethal thresholds), metabolic rates (routine, maximum, and minimum metabolic rate, aerobic scope), enzyme reaction rates and concentrations, oxidative stress and consequences (e.g., DNA damage, teleomere attrition), performance measurements (e.g., heart rates, sprint speed, swim speed, growth rates), body composition (e.g., fatty acid profiles) and immune traits (e.g., cell function, antibodies). Studies were also included if they measured the relationship between physiological traits and body size (e.g., metabolic scaling), reproductive allocation, or survival across taxa. Studies on behavior were only included if the synthesis was explicitly interested in the proximate physiological drivers of behavioral variation. We also included meta-analyses that quantified plasticity in physiological traits in response to environmental variation. While many studies in comparative physiology collect data as part of a comparative analysis, we focus here only on studies collating data from existing literature because this is the major data source in quantitative syntheses. Quantitative syntheses interested in phenology, bio-mechanics, sexual size dimorphism (i.e., SSD, except if the study was interested in sexual differences in physiological traits), those focused on validating physiological methodology, or characterising population genetic structure, gene transcription and quantitative genetics were excluded. There were also meta-analyses that focused on colour evolution and contrast which varied in their goals but were most often not focused on physiological questions. As such, these studies were also excluded. We restricted our focus to animal studies because most plant synthesis papers were focused on ecological (e.g., quantified changes in abundance and distribution of species) or agricultural questions (e.g., studies on soil microbial function, nitrogen and phosphorus capture, methane output and greenhouse gas emissions).

@@

Quantitative syntheses in comparative physiology commonly analyse raw summary statistics (`r with(effects, effects[effect == "Means","prop"])*100`% of studies), followed closely by Fishers transformed correlation coefficients (Zr) (`r with(effects, effects[effect == "Zr","prop"])*100`% of studies), standardised mean differences (SMD) and log response ratios (lnRR) (`r with(effects, effects[effect == "SMD","prop"])*100`% and `r with(effects, effects[effect == "lnRR","prop"])*100`%, respectively; Figure \@ref(fig:figsurv)A). In many cases, studies make use of multiple effect size measures. Approximately `r with(hetero, hetero[cont_moderator_variable == "yes","percent"])*100`% of effect sizes used within the study we deemed to be impacted by 'nuisance heterogeneity' in some form or another (Figure \@ref(fig:figsurv)B), but of the studies where this was the case, only `r with(hetero_dealt, hetero_dealt[cont_moderator_variable_modelled == "yes","percent"])*100`% of the studies clearly dealt with it (Figure \@ref(fig:figsurv)B); often through meta-regression modelling. Overall, `r with(weighted, weighted[meta_analysis_weighted == "weighted model","percent"])*100`% of studies used a weighted meta-analytic model, however, of the studies self described as 'meta-analyses', `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","percent"])*100`% (N = `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","n"])` of `r with(weighted_meta, sum(weighted_meta[,"n"]))`) did not account for sampling variance. Many studies used 'Linear Regression' (including linear and generalised linear mixed effects models) as the main type of model fit to the data (Figure \@ref(fig:figsurv)C). Of the studies using weighted meta-analytic models `r with(het_reported, sum(het_reported[1:2,"percent"])*100)`% (N = `r with(het_reported, sum(het_reported[1:2,"n"]))` meta-analyses) did not report any measure of heterogeneity, which is unfortunate given how important these measures are to interpreting effects.

@@

Meta-analysis requires comparative physiologists to make use of effect sizes that are suitable for answering their specific question. The key features of all effect sizes used in meta-analyses are that: 1) they are statistical parameters that have been placed on a common scale so that they can be compared across studies and 2) they have some associated measure of precision (i.e., sampling variance) [@Koricheva2013; @Borenstein2009; @Schmid2021]. Effect sizes in comparative physiology are quite variable; they can be the mean difference between two groups [e.g., difference in mean between a control and treatment group; e.g., @WuSeebacher2020], the slope of a regression (e.g., between body size and metabolic rate), a correlation between two variables (e.g., between hormone levels and behaviour), or even the raw mean (or variance) itself [e.g., comparing CTmax or CTmin across species - e.g., @Sunday2014]. In many cases, different effect sizes might be used depending on the question. For example, to investigate the effect of temperature on a physiological trait, one can compile studies experimentally manipulating temperature and analyse the mean 'effect' by creating contrast-based effect sizes, or alternatively, just model the trait in each temperature treatment. The choice as to which one to use boils down to the types of data available from primary studies, the specific question of interest, and whether studies are observational or experimental in nature. Considering our temperature example above, if we are interested in the magnitude of change between temperature treatments, we may want to use a contrast-based effect size, such as the log response ratio $\ln{RR}$. Instead, if we are interested in how the mean of some physiological trait changes with temperature, we may model the mean itself. In the latter scenario, as long as all the means are measured in the same units (e.g., Celsius; or can be converted to the same units) then the mean may provide the desired answer to the original question. We provide details on different effect measures and associated sample sizes in Table \@ref(tab:tablea).

@@

# I'm just listing a bunch here, but we should think about what we want to include. Here's an example of how we can make a table with all equations. Fun! Really useful website here: https://math.meta.stackexchange.com/questions/5020/mathjax-basic-tutorial-and-quick-reference. Something wrong with character coding when rendering word document. Has to do with the log odds ratio equation. Renders fine in html, but just stops when word
names <- c("Mean",
           "Log Standard deviation, lnSD",
           "Log Response Ratio, lnRR",
           # leave this for now??? - we introduce this later
           "Standardised Mean Difference, SMD",
           "Zr (Fisher Transformation of \n correlation Coefficient, r)")
           #"Log Odds Ratio",
           #"")

@@

As can be seen in Table \@ref(tab:tablea), none of the effect measures control for 'nuisance heterogeneity'. For example, the magnitude of lnRR or SMD might depend on just how different the temperatures, dosages, pH or water potential applied in each treatment. The same applies for modelling the means. For example, standard metabolic rate (SMR) is temperature dependent. As such, one needs to standardise the temperature measurement to compare across studies [@White2006; @Uyeda2017] (if, of course, temperature is not of interest in of itself). While we could model this explicitly, this may not always be desirable. It can often be useful to standardise the effect to both ease its interpretation and simplify modelling. In Box 1 we provide readers with guidance on how comparative physiologists can apply corrections to the effect measure itself. We highlight two examples in the box focusing mainly on generality. Below, we show how to formalize a traditional effect size, $lnRR$, into a common measure of interest to comparative physiologists, $Q_{10}$, and show how this new effect size can capture “nuisance heterogeneity”.

@@

A common experiment in comparative physiology is to manipulate the temperature organisms experience and measure some physiological rate (e.g., metabolic rate). Using the effect size measures in Table \@ref(tab:tablea) would result in the effect sizes varying with the temperatures applied within and across across studies contributing to 'comparability' issues. For example, an effect size for a temperature manipulation of 10°C would be larger than an effect size for a temperature manipulation of 5°C (see Figure \@ref(fig:fig1)). As such, it is common to standardize effects by temperature. One common effect size measure already used in comparative physiology to compare physiological rates is the temperature coefficient $Q_{10}$ [e.g., @Rodgers2021; @Seebacher2015; @Havird2020]. This effect size describes the multiplicative change in physiological rates across a 10°C temperature change. Higher $Q_{10}$ values indicate larger changes in physiological rates. However, currently there is no formal sampling variance associated with $Q_{10}$ in the literature, making it challenging to make use of the full power of weighted meta-analytic models. Interestingly, $Q_{10}$ can be seen as a variant of *lnRR*, meaning that we can derive a $Q_{10}$- based effect size and sampling error using the well-known mathematical properties of *lnRR*. This opens up the meta-analytic toolbox and improves upon our ability to account for well known sources of non-independence that are typical in meta-analysis [@Lajeunesse2011; @Noble2017].

Prior to showing how the relevant $Q_{10}$ effect size can be calculated it is useful to understand its similarities to *lnRR*. The *lnRR* described by @Hedges1999 and extended by @Lajeunesse2015 and can be calculated as follows [but see also @Senior2020]:

@@

In equation \@ref(eq:lnRR), $M_{1}$ is the mean of group 1 (e.g., a control group), where as $M_{2}$ is the mean of group 2 (e.g., a treatment group). The mean, $M_{i}$, for group *i* can be any measurement type (e.g. a physiological rate, mass etc) so long as the measurement variable is ratio scale. Natural log transformation of this ratio makes the effect size normally distributed which is commonly assumed by meta-analytic models. Equation \@ref(eq:slnRR) is the analytical solution for *lnRR*'s sampling variance where, $SD_{1}^2$ and $SD_{2}^2$ is the sample variance in group 1 and 2, respectively and $N_{1}$ and $N_{2}$ is the sample size in group 1 and 2.

@@

Equation \@ref(eq:lnq10) is essentially a temperature corrected equivalent of lnRR when the numerator and denominator are measured at different temperatures. This allows one to compare the mean of two temperature treatments directly regardless of the temperatures that these groups have been measured. Here, we will refer to this as the log $Q_{10}$ response ratio, $lnRR_{Q_{10}}$. Notably, using this effect rather than the traditional $Q_{10}$ has two statistical advantages: 1) zero becomes biologically meaningful as 0 means two rates are the exactly the same and 'significantly different from zero' means that two rates are significantly different to each other and 2) $lnRR_{Q_{10}}$ is more likely to satisfy the assumption of residual normality than $Q_{10}$ (see @Hedges1999). The recognition of this equivalence means that we can calculate the sampling variance for equation \@ref(eq:lnq10) as follows:

@@

Our example using $Q_{10}$, a well-known effect measure in comparative physiology, shows how common effect size's can be re-formalised to account for 'nuisance heterogeneity', in this case temperature. In doing so, we are making assumptions about the nature of temperature effects on an effect size. Nonetheless, we can apply similar approaches to other commonly used effect size metrics. We describe more generally how this can be done in Box 1, applying similar principles to standardise temperature differences for slopes and standardised mean differences.

@@

A common goal of meta-analysis is to obtain an overall meta-analytic mean effect size estimate and some form of uncertainty around that mean, like a 95% confidence interval (CI). Including a prediction interval (PI) can also be useful (discussed below) [@Nakagawa2021c]. Estimating an overall mean and 95% CI is generally done with a simple random effects or multilevel meta-analytic model that accounts for effect size sampling error [@Nakagawa2017; @NakagawaSantos2012]. More often than not, meta-analyses in comparative physiology will have complex hierarchical structure. For example, multiple effect sizes may be taken from a single study on traits measured on the same individuals, or come from many different species that share an evolutionary history [@Cinar2021; @Noble2017]. As such, it is far more likely that comparative physiologists will need to apply a multilevel meta-analytic model to control for the various sources of non-independence, and whenever possible, they should [@Cinar2021; @NakagawaSantos2012; @Song2021; @Nakagawa2021; @Noble2017]. Multilevel models are also extremely useful in describing the various sources contributing to effect size variance (a feature we describe below). For the purpose of our discussion, we will assume one needs to use more advanced multilevel models. Having said that, in some instances, it may be sufficient for a meta-analysis to make use of traditional random effects models. We refer readers to excellent reviews and books that overview the distinctions between common (fixed) effect, random effect and multilevel models and when one might want (or not want) to apply these models [@Nakagawa2017; @NakagawaSantos2012; @Koricheva2013; @Borenstein2009; @Schmid2021].

@@

where $Y_{i}$ is the *i*th effect size (*i* = 1 ... $N_{ES}$, the number of effect sizes), $\beta_{0}$ is the overall meta-analytic mean, $a_{k[i]}$ is the phylogenetic effect (a random effect) for species *k* applied to effect size *i* [i.e., species effects because of shared evolutionary history, @Cinar2021]. Phylogenetic effects are assumed to be normally distributed deviates sampled from a distribution with a mean of 0 and variance $\sigma^{2}_{phylogeny}$ [i.e., *N*$(0, \sigma^{2}_{phylogeny}\textbf{A})$], where $\textbf{A}$ is a phylogenetic correlation matrix derived from a phylogenetic tree. $sp_{k[i]}$ is an additional species-specific effect for the *k*th species applied to effect size *i* [i.e., these are species-specific effects because of shared ecology and other factors, @Cinar2021], $s_{j[i]}$ is the study-specific effect for the *j*th study applied to effect size *i*, $e_{i}$ is the effect size-specific effect (or within study effect, or residuals) for the *i*th effect size, and $m_{i}$ is the sampling variance effect around the *i*th effect as a result of variation in precision among each effect sizes, $v_{i}$, which is known as the sampling variance for the effect. This model is a 'weighted' multilevel meta-analysis model because the estimation of the overall meta-analytic mean, $\beta_{0}$, is partially weighted by the inverse sampling variance of each effect size. Importantly, if we have effect sizes that are derived from the *same* sample of organisms (e.g., because traits are measured on the same sets of individuals) then we need to remove this dependency from the calculation of each $v_{i}$; we can then model the known sampling covariance between effect sizes by modifying $\textbf{I}$ to include off-diagonals that account for this covariance [See Box 1 and @Noble2017]. The fact that we can obtain an overall estimate of sampling variance (i.e., $\sigma^2_{m}$, see below for how it is defined) is very important because it provides a way for us to understand just how much variation in effects are the result of sampling differences across studies versus other, potentially important sources of variation such as differences in the methods applied, species differences or even real biological effects that impact the effect size a study is likely to observe [@Nakagawa2017; @NakagawaSantos2012]. This is formalised in what we call 'heterogeneity analysis' which we describe below. Importantly, this model provides us with the answer to our overall question: Across all the species and studies manipulating egg corticosterone, how much of an overall effect does corticosterone have on reproduction relative to control conditions?

Of course, the overall effect does not provide us with a way to interpret its magnitude in the context of how much CORT has been applied in the treatments used in the sample of studies. The most we can say about it, is that the effect is small, moderate or large out of the sample of effect sizes extracted from this set of studies. In the next section, we show how to 'correct' the overall effect size so that its interpretation is more meaningful.

>>>>

    # If you have issues with flextable and equatags, make sure you have node.js and use mathjax_install(). See here:https://github.com/ardata-fr/equatags
  pacman::p_load(flextable, tidyverse, equatags, png, magick)  
```

@@

Meta-analysis is always concerned with effect heterogeneity; in other words, the factors that drive differences in the direction and magnitude of effects within and across studies [@Nakagawa2017; @Gurevitch2018; @Cooperetal2019; @Borenstein2019]. The concept of heterogeneity is vitally important because it tells us how general our findings are likely to be and how much of the variance we see is the result of real biology or methodological differences (not just sampling variance) [@Nakagawa2017; @Borenstein2019]. There are many phenomena driving effect heterogeneity, some of which are interesting and others that are less so.

@@

```{r, fig1, fig.cap="*Sources of effect size heterogeneity in meta-analyses*. Heterogeneity can come in various forms, from effect measures varying as a result of shared evolutionary history (i.e., phylogeny), different trait types measured, and sources of 'nuisance heterogeneity'. In this example, nuisance heterogeneity is generated by differences in the experimental temperatures at which the animals were tested. However, nuisance heterogeneity can come in various other forms such as differences in dosage, or biochemical variables between used in the experiments. We can control for these types of heterogeniety by reformalising effect sizes and/or using multilevel metaregression as we describe in our review."}
image <- image_read(here::here("figs", "Fig1.png"))
image
```

@@

We tend to agree with @Glass2015, and his sentiment is of particular relevance to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms, including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are different. By synthesizing diverse study outcomes, not only does it allow to report on overall effects (if the question is relevant to ask), but it allows us to also understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction), and approaches that can explain the diversity of effect outcomes observed. Doing so provides a richer set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included in a meta-analysis, and how this can affect interpretations made -- an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. If the estimate of an overall effect coincides with low heterogeneity estimates then the effects are reasonably consistent [@Borenstein2019]. If, however, an overall effect is accompanied by large heterogeneity estimates, then we would shift focus and explore drivers of that heterogeneity. For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant and may even be misleading. Therefore, understanding overall effects in each of these groups separately, and assess the extent to which they differ, may be more useful.

@@

Before embarking overviewing strategies to deal with 'nuisance heterogeneity', it is worth first understanding how comparative physiologists are currently conducting meta-analyses and other quantitative syntheses. As discussed above, different types of questions and experimental approaches will mix 'apples and oranges' to varying extents, but this heterogeneity (or basket of fruit) is often the main interest to comparative biologists. We conducted a literature survey to better understand the different types of effect sizes being used, their susceptibility to known sources of methodological / environmental variability across studies, and how these sources are controlled, we conducted a literature survey. To accomplish this goal, we searched for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to continuous variation in how treatments were applied (e.g., for contrast-based effects) or the conditions under which traits were measured (i.e., for correlations) across studies (i.e., 'nuisance heterogeneity')? 3) If so, was the variation accounted for in the effect size or during analysis? and finally, 4) how many analyses used 'weighted' meta-analytic models?

@@

We included comparative / meta-analyses that were focused on comparative physiological questions. Narrative reviews and comments were not considered because they did not collate quantitative data from the literature. Studies needed to have collected physiological trait data from the literature such as hormones, thermal physiology (CTmax, CTmin, thermal preferences, thermal breadth, LT50, lethal thresholds), metabolic rates (routine, maximum, and minimum metabolic rate, aerobic scope), enzyme reaction rates and concentrations, oxidative stress and consequences (e.g., DNA damage, teleomere attrition), performance measurements (e.g., heart rates, sprint speed, swim speed, growth rates), body composition (e.g., fatty acid profiles) and immune traits (e.g., cell function, antibodies). Studies were also included if they measured the relationship between physiological traits and body size (e.g., metabolic scaling), reproductive allocation, or survival across taxa. Studies on behavior were only included if the synthesis was explicitly interested in the proximate physiological drivers of behavioral variation. We also included meta-analyses that quantified plasticity in physiological traits in response to environmental variation. While many studies in comparative physiology collect data as part of a comparative analysis, we focus here only on studies collating data from existing literature because this is the major data source in quantitative syntheses. Quantitative syntheses interested in phenology, bio-mechanics, sexual size dimorphism (i.e., SSD, except if the study was interested in sexual differences in physiological traits), those focused on validating physiological methodology, or characterising population genetic structure, gene transcription, and quantitative genetics were excluded. There were also meta-analyses that focused on colour evolution and contrast which varied in their goals but were most often not focused on physiological questions. As such, these studies were also excluded. We restricted our focus to animal studies because most plant synthesis papers were focused on ecological (e.g., quantified changes in abundance and distribution of species) or agricultural questions (e.g., studies on soil microbial function, nitrogen and phosphorus capture, methane output, and greenhouse gas emissions).

@@

Quantitative syntheses in comparative physiology commonly analyse raw summary statistics (`r with(effects, effects[effect == "Means","prop"])*100`% of studies), followed closely by Fishers transformed correlation coefficients (Zr) (`r with(effects, effects[effect == "Zr","prop"])*100`% of studies), standardised mean differences (SMD) (`r with(effects, effects[effect == "SMD","prop"])*100`%  of studies), and log response ratios (lnRR) (`r with(effects, effects[effect == "lnRR","prop"])*100`%, of studies) (Figure \@ref(fig:figsurv)A). In many cases, studies make use of multiple effect size measures. Approximately `r with(hetero, hetero[cont_moderator_variable == "yes","percent"])*100`% of effect sizes used within the study we deemed to be impacted by 'nuisance heterogeneity' in some form or another (Figure \@ref(fig:figsurv)B), but of the studies where this was the case, only `r with(hetero_dealt, hetero_dealt[cont_moderator_variable_modelled == "yes","percent"])*100`% of the studies clearly dealt with it (Figure \@ref(fig:figsurv)B), often through meta-regression modelling. Overall, `r with(weighted, weighted[meta_analysis_weighted == "weighted model","percent"])*100`% of studies used a weighted meta-analytic model; however, of the studies self described as 'meta-analyses', `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","percent"])*100`% (N = `r with(weighted_meta, weighted_meta[meta_analysis_weighted == "unweighted model","n"])` of `r with(weighted_meta, sum(weighted_meta[,"n"]))`) did not account for sampling variance. Many studies used 'Linear Regression' (including linear and generalised linear mixed effects models) as the main type of model fit to the data (Figure \@ref(fig:figsurv)C). Of the studies using weighted meta-analytic models `r with(het_reported, sum(het_reported[1:2,"percent"])*100)`% (N = `r with(het_reported, sum(het_reported[1:2,"n"]))` meta-analyses) did not report any measure of heterogeneity, which is unfortunate given how important these measures are to interpreting effects.

@@

Meta-analysis requires comparative physiologists to make use of effect sizes that are suitable for answering their specific question. The key features of all effect sizes used in meta-analyses are that: 1) they are statistical parameters that have been placed on a common scale so that they can be compared across studies and 2) they have some associated measure of precision (i.e., sampling variance) [@Koricheva2013; @Borenstein2009; @Schmid2021]. Effect sizes in comparative physiology are quite variable; they can be the mean difference between two groups [e.g., difference in mean between a control and treatment group; e.g., @WuSeebacher2020], the slope of a regression (e.g., between body size and metabolic rate), a correlation between two variables (e.g., between hormone levels and behaviour), or even the raw mean (or variance) itself [e.g., comparing CTmax or CTmin across species - e.g., @Sunday2014]. In many cases, different effect sizes might be used depending on the question. For example, to investigate the effect of temperature on a physiological trait, one can compile studies experimentally manipulating temperature and analyse the mean 'effect' by creating contrast-based effect sizes, or alternatively, just model the trait in each temperature treatment. The choice as to which one to use boils down to the types of data available from primary studies, the specific question of interest, and whether studies are observational or experimental in nature. Considering our temperature example above, if we are interested in the magnitude of change between temperature treatments, we may want to use a contrast-based effect size, such as the log response ratio $\ln{RR}$. Instead, if we are interested in how the mean of some physiological trait changes with temperature, we may model the mean itself. In the latter scenario, as long as all the means are measured in the same units (e.g., Celsius) or can be converted to the same units, then the mean may provide the desired answer to the original question. We provide details on different effect measures and associated sample sizes in Table \@ref(tab:tablea).

@@

# I'm just listing a bunch here, but we should think about what we want to include. Here's an example of how we can make a table with all equations. Fun! Really useful website here: https://math.meta.stackexchange.com/questions/5020/mathjax-basic-tutorial-and-quick-reference. Something wrong with character coding when rendering word document. Has to do with the log odds ratio equation. Renders fine in html, but just stops when word
names <- c("Mean",
           "Log Standard deviation, lnSD",
           "Log Response Ratio, lnRR",
           # leave this for now??? - we introduce this later
           "Standardised Mean Difference, SMD",
           "Zr (Fisher Transformation of \n Correlation Coefficient, r)")
           #"Log Odds Ratio",
           #"")

@@

As can be seen in Table \@ref(tab:tablea), none of the effect measures control for 'nuisance heterogeneity'. For example, the magnitude of lnRR or SMD might depend on differences in the temperatures, dosages, pH, or water potential applied to each treatment. The same applies for modelling the means. For example, standard metabolic rate (SMR) is temperature dependent. As such, one needs to standardise the temperature measurement to compare across studies [@White2006; @Uyeda2017] (if, of course, temperature is not of interest in of itself). While we could model this explicitly, this may not always be desirable. It can often be useful to standardise the effect to both ease its interpretation and simplify modelling. In Box 1 we provide readers with guidance on how comparative physiologists can apply corrections to the effect measure itself. We highlight two examples in the box focusing mainly on generality. Below, we show how we can formalise a traditional effect size, such as $lnRR$, to capture this ‘nuisance variation’ using a common, well understood effect measure of interest to comparative physiologists ($Q_{10}$).

@@

A common experiment in comparative physiology is to manipulate the temperature organisms experience and measure some physiological rate (e.g., metabolic rate). Using the effect size measures in Table \@ref(tab:tablea) would result in the effect sizes varying with the temperatures applied within and across studies contributing to 'comparability' issues. For example, an effect size for a temperature manipulation of 10°C would be larger than an effect size for a temperature manipulation of 5°C (see Figure \@ref(fig:fig1)). As such, it is common to standardize effects by temperature. One common effect size measure already used in comparative physiology to compare physiological rates is the temperature coefficient $Q_{10}$ [e.g., @Rodgers2021; @Seebacher2015; @Havird2020]. This effect size describes the multiplicative change in physiological rates across a 10°C temperature change. Higher $Q_{10}$ values indicate larger changes in physiological rates. However, currently there is no formal sampling variance associated with $Q_{10}$ in the literature, making it challenging to make use of the full power of weighted meta-analytic models. Interestingly, $Q_{10}$ can be seen as a variant of *lnRR*, meaning that we can derive a $Q_{10}$- based effect size and sampling error using the well-known mathematical properties of *lnRR*. This opens up the meta-analytic toolbox and improves upon our ability to account for well known sources of non-independence that are typical in meta-analysis [@Lajeunesse2011; @Noble2017].

Prior to showing how the relevant $Q_{10}$ effect size can be calculated it is useful to understand its similarities to *lnRR*. The *lnRR* described by @Hedges1999 and extended by @Lajeunesse2015 can be calculated as follows [but see also @Senior2020]:

@@

In equation \@ref(eq:lnRR), $M_{1}$ is the mean of group 1 (e.g., a control group), where as $M_{2}$ is the mean of group 2 (e.g., a treatment group). The mean, $M_{i}$, for group *i* can be any measurement type (e.g. a physiological rate, mass etc) so long as the measurement variable is ratio scale. Natural log transformation of this ratio makes the effect size normally distributed which is commonly assumed by meta-analytic models. Equation \@ref(eq:slnRR) is the analytical solution for *lnRR*'s sampling variance where $SD_{1}^2$ and $SD_{2}^2$ is the sample variance in group 1 and 2, respectively and $N_{1}$ and $N_{2}$ is the sample size in group 1 and 2.

@@

Equation \@ref(eq:lnq10) is essentially a temperature corrected equivalent of lnRR when the numerator and denominator are measured at different temperatures. This allows one to compare the mean of two temperature treatments directly regardless of the temperatures at which these groups have been measured. Here, we will refer to this as the log $Q_{10}$ response ratio, $lnRR_{Q_{10}}$. Notably, using this effect rather than the traditional $Q_{10}$ has two statistical advantages: 1) zero becomes biologically meaningful as 0 means two rates are the exactly the same and 'significantly different from zero' means that two rates are significantly different to each other and 2) $lnRR_{Q_{10}}$ is more likely to satisfy the assumption of residual normality than $Q_{10}$ (see @Hedges1999). The recognition of this equivalence means that we can calculate the sampling variance for equation \@ref(eq:lnq10) as follows:

@@

Our example using $Q_{10}$, a well-known effect measure in comparative physiology, shows how common effect size's can be re-formalised to account for 'nuisance heterogeneity' (in this case temperature). In doing so, we are making assumptions about the nature of temperature effects on an effect size. Nonetheless, we can apply similar approaches to other commonly used effect size metrics. We describe more generally how this can be done in Box 1, applying similar principles to standardise temperature differences for slopes and standardised mean differences.

@@

A common goal of meta-analysis is to obtain an overall meta-analytic mean effect size estimate and some form of uncertainty around that mean, like a 95% confidence interval (CI). Including a prediction interval (PI) can also be useful (discussed below) [@Nakagawa2021c]. Estimating an overall mean and 95% CI is generally done with a simple random effects or multilevel meta-analytic model that accounts for effect size sampling error [@Nakagawa2017; @NakagawaSantos2012]. More often than not, meta-analyses in comparative physiology will have complex hierarchical structure. For example, multiple effect sizes may be taken from a single study on traits measured on the same individuals or from many different species that share an evolutionary history [@Cinar2021; @Noble2017]. As such, it is far more likely that comparative physiologists will need to apply a multilevel meta-analytic model to control for the various sources of non-independence, and, whenever possible, they should [@Cinar2021; @NakagawaSantos2012; @Song2021; @Nakagawa2021; @Noble2017]. Multilevel models are also extremely useful in describing the various sources contributing to effect size variance (a feature we describe below). For the purpose of our discussion, we will assume one needs to use more advanced multilevel models. Having said that, in some instances, it may be sufficient for a meta-analysis to make use of traditional random effects models. We refer readers to excellent reviews and books that overview the distinctions between common (fixed) effect, random effect and multilevel models and when one might want (or not want) to apply these models [@Nakagawa2017; @NakagawaSantos2012; @Koricheva2013; @Borenstein2009; @Schmid2021].

@@

where $Y_{i}$ is the *i*th effect size (*i* = 1 ... $N_{ES}$, the number of effect sizes), $\beta_{0}$ is the overall meta-analytic mean, $a_{k[i]}$ is the phylogenetic effect (a random effect) for species *k* applied to effect size *i* [i.e., species effects because of shared evolutionary history, @Cinar2021]. Phylogenetic effects are assumed to be normally distributed deviates sampled from a distribution with a mean of 0 and variance $\sigma^{2}_{phylogeny}$ [i.e., *N*$(0, \sigma^{2}_{phylogeny}\textbf{A})$], where $\textbf{A}$ is a phylogenetic correlation matrix derived from a phylogenetic tree. $sp_{k[i]}$ is an additional species-specific effect for the *k*th species applied to effect size *i* [i.e., these are species-specific effects because of shared ecology and other factors, @Cinar2021], $s_{j[i]}$ is the study-specific effect for the *j*th study applied to effect size *i*, $e_{i}$ is the effect size-specific effect (or within study effect, or residuals) for the *i*th effect size, and $m_{i}$ is the sampling variance effect around the *i*th effect as a result of variation in precision among each effect size, $v_{i}$ (which is known as the sampling variance for the effect). This model is a 'weighted' multilevel meta-analysis model because the estimation of the overall meta-analytic mean, $\beta_{0}$, is partially weighted by the inverse sampling variance of each effect size. Importantly, if we have effect sizes that are derived from the *same* sample of organisms (e.g., because traits are measured on the same sets of individuals), then we need to remove this dependency from the calculation of each $v_{i}$; we can then model the known sampling covariance between effect sizes by modifying $\textbf{I}$ to include off-diagonals that account for this covariance [See Box 1 and @Noble2017]. The fact that we can obtain an overall estimate of sampling variance (i.e., $\sigma^2_{m}$, see below for how it is defined) is very important because it provides a way for us to understand just how much variation in effects are the result of sampling differences across studies versus other, potentially important sources of variation such as differences in the methods applied, species differences, or even real biological effects that impact the effect size a study is likely to observe [@Nakagawa2017; @NakagawaSantos2012]. This is formalised in what we call 'heterogeneity analysis' which we describe below. Importantly, this model provides us with the answer to our overall question: Across all the species and studies manipulating egg corticosterone, how much of an overall effect does corticosterone have on reproduction relative to control conditions?

Of course, the overall effect does not provide us with a way to interpret its magnitude in the context of how much CORT has been applied in the treatments used in the sample of studies. The most we can say about it is that the effect is small, moderate or large out of the sample of effect sizes extracted from this set of studies. In the next section, we show how to 'correct' the overall effect size so that its interpretation is more meaningful.
